---
title: "titanic"
author: "ianalexmac"
date: "1/26/2022"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
### install packages
```{r}
#install.packages('rattle')
#install.packages('rpart.plot')
#install.packages('RColorBrewer')

```



### libraries
```{r}
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)

```

### import
```{r}
test  <- read.csv("~/R/titanic/test.csv")
train <- read.csv("~/R/titanic/train.csv")

```

### explore
```{r}
summary(test)
summary(train)

str(test)
str(train)

prop.table(table(train$Survived))
```

### test$Fare NA (row 153)
add using median fare of 3rd class passengers
```{r}
class.3.fare <- median(test$Fare[test$Pclass == 3], na.rm = TRUE)
test$Fare <- ifelse(is.na(test$Fare), class.3.fare, test$Fare)

summary(test)
```

### clean NA's from Age
lots of NA's in age (86 in test, 177 in train)
make seperate data frames omitting Age rows with NA's
```{r}
a <- which(is.na(test$Age))
test.no.na <- test[-a,]
sum(is.na(test.no.na))

b <- which(is.na(train))
train.no.na <- train[-b,]
sum(is.na(test.no.na))

```

### explore proportion of male/female survivors
```{r}
prop.table(table(train$Sex, train$Survived), 1)

```


### create submission dataframe for Kaggle
prediction: females live, males die
.7655 accuracy on Kaggle!
```{r}
test$Survived <- 0
test$Survived[test$Sex == 'female'] <- 1

submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)

write.csv(submit, file = "females.survive.csv", row.names = FALSE)

```

### add column to bin fare cost
```{r}
train$FareCost <- '$$$$'
train$FareCost[train$Fare < 30 & train$Fare >= 20] <- '$$$'
train$FareCost[train$Fare < 20 & train$Fare >= 10] <- '$$'
train$FareCost[train$Fare < 10] <- '$'

```


###more exploration
```{r}
aggregate(Survived ~ Pclass + Sex, data = train, FUN = function(x) {sum(x)/length(x)})
aggregate(Survived ~ Pclass + Sex + Embarked, data = train, FUN = function(x) {sum(x)/length(x)})
aggregate(Survived ~ Pclass + FareCost + Sex, data = train, FUN = function(x) {sum(x)/length(x)})

```
3rd class females with expensive tickets perished? ok


### make another submission
prediction: females survived, except 3rd class females with expensive tickets
Kaggle: .7799 accuracy
```{r}
test$Survived <- 0
test$Survived[test$Sex == 'female'] <- 1
test$Survived[test$Sex == 'female' & test$Pclass == 3 & test$Fare >= 20] <- 0

submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)
write.csv(submit, file = "exp.3rd.f.perish.csv", row.names = FALSE)

```


### decision tree
write submission
Kaggle: .78469 accurate
```{r}
fit <- rpart(Survived ~ Pclass + Sex + Fare + Age + Parch + Embarked, data = train, method = "class")

fancyRpartPlot(fit)

prediction <- predict(fit, test, type = "class")
submit <- data.frame(PassengerId = test$PassengerId, Survived = prediction)
write.csv(submit, file = "tree.csv", row.names = FALSE)




```