---
title: "titanic"
author: "ianalexmac"
date: "1/26/2022"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
### install packages
```{r}
#install.packages('rattle')
#install.packages('rpart.plot')
#install.packages('RColorBrewer')
#install.packages("randomForest")
#install.packages("party")


```



### libraries
```{r}
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(randomForest)
library(party)

```

### import
```{r}
test  <- read.csv("~/R/titanic/test.csv")
train <- read.csv("~/R/titanic/train.csv")

```

### explore
```{r}
summary(test)
summary(train)

str(test)
str(train)

prop.table(table(train$Survived))
```

### test$Fare NA (row 153)
add using median fare of 3rd class passengers
```{r}
class.3.fare <- median(test$Fare[test$Pclass == 3], na.rm = TRUE)
test$Fare <- ifelse(is.na(test$Fare), class.3.fare, test$Fare)

summary(test)

train$Sex <- as.factor(train$Sex)
test$Sex <- as.factor(test$Sex)

```

### clean NA's from Age
lots of NA's in age (86 in test, 177 in train)
make seperate data frames omitting Age rows with NA's
```{r}
a <- which(is.na(test$Age))
test.no.na <- test[-a,]
sum(is.na(test.no.na))

b <- which(is.na(train))
train.no.na <- train[-b,]
sum(is.na(test.no.na))

```

### explore proportion of male/female survivors
```{r}
prop.table(table(train$Sex, train$Survived), 1)

```


### create submission dataframe for Kaggle
prediction: females live, males die
.7655 accuracy on Kaggle!
```{r}
test$Survived <- 0
test$Survived[test$Sex == 'female'] <- 1

submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)

write.csv(submit, file = "females.survive.csv", row.names = FALSE)

```

### add column to bin fare cost
```{r}
train$FareCost <- '$$$$'
train$FareCost[train$Fare < 30 & train$Fare >= 20] <- '$$$'
train$FareCost[train$Fare < 20 & train$Fare >= 10] <- '$$'
train$FareCost[train$Fare < 10] <- '$'

test$FareCost <- '$$$$'
test$FareCost[test$Fare < 30 & test$Fare >= 20] <- '$$$'
test$FareCost[test$Fare < 20 & test$Fare >= 10] <- '$$'
test$FareCost[test$Fare < 10] <- '$'

```


###more exploration
```{r}
aggregate(Survived ~ Pclass + Sex, data = train, FUN = function(x) {sum(x)/length(x)})
aggregate(Survived ~ Pclass + Sex + Embarked, data = train, FUN = function(x) {sum(x)/length(x)})
aggregate(Survived ~ Pclass + FareCost + Sex, data = train, FUN = function(x) {sum(x)/length(x)})

```
3rd class females with expensive tickets perished? ok


### make another submission
prediction: females survived, except 3rd class females with expensive tickets
Kaggle: .7727  accuracy
```{r}
test$Survived <- 0
test$Survived[test$Sex == 'female'] <- 1
test$Survived[test$Sex == 'female' & test$Pclass == 3 & test$Fare >= 20] <- 0

submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)
write.csv(submit, file = "exp.3rd.f.perish.csv", row.names = FALSE)

```


### decision tree
write submission
Kaggle: .7703 accuracy
```{r}
fit <- rpart(Survived ~ Pclass + Sex + Fare + Age + Parch + Embarked, data = train, method = "class")

fancyRpartPlot(fit)

prediction <- predict(fit, test, type = "class")
submit <- data.frame(PassengerId = test$PassengerId, Survived = prediction)
write.csv(submit, file = "tree.csv", row.names = FALSE)


```

### random forest preparation
cannot handle NA's, have 263 in Age
use decision tree to fill values
```{r}
test$Survived <- NA
full <- rbind(train, test)
full$Name <- as.character(full$Name)

summary(full)

age.predict <- rpart(Age ~ Pclass + Sex + Parch + Fare + Embarked,
                     data = full[!is.na(full$Age),],
                     method = "anova")

full$Age[is.na(full$Age)] <- predict(age.predict, full[is.na(full$Age),])

summary(full)

```


### clean blanks in Embarked
```{r}
full$Embarked <- as.factor(full$Embarked)
summary(full$Embarked)
which(full$Embarked == '')
full$Embarked[c(62, 830)] = "S"
summary(full$Embarked)

```


###split full back into train and test
```{r}
train <- full[1:891,]
test <- full[892:1309,]


```

### random forest
```{r}
set.seed(666)

fit <- randomForest(as.factor(Survived) ~ Pclass + Sex + Age + Fare + Embarked,
                    data = train,
                    importance = TRUE,
                    ntree = 2000)

varImpPlot(fit)
```


### random forest submission
Kaggle: .7823 accuracy! 
```{r}
prediction <- predict(fit, test)
submit <- data.frame(PassengerId = test$PassengerId, Survived = prediction)
write.csv(submit, file = "forest.csv", row.names = FALSE)

```


### conditional inference trees (party mode)
```{r}
library(party)
set.seed(666)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked,
               data = train,
               controls = cforest_unbiased(ntree = 2000, mtry = 3))

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare +
                                       Embarked,
                 data = train, 
                 controls = cforest_unbiased(ntree = 2000, mtry = 3))

```

### conditional forest submission
Kaggle: .7847 accurate
```{r}
prediction <- predict(fit, newdata = test, OOB=TRUE, type = "response")

submit <- data.frame(PassengerId = test$PassengerId, Survived = prediction)
write.csv(submit, file = "cond.forest.csv", row.names = FALSE)


```
